---
title: "Multi-series macro panels: fetch, transform, widen, plot"
author: "Charles Coverdale"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-series macro panels: fetch, transform, widen, plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
key_available <- nzchar(Sys.getenv("FRED_API_KEY"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = key_available
)
```

This vignette walks through the most common starting workflow with `fred`:
fetching several series in one call, applying a server-side transformation,
pivoting to wide format for cross-series analysis, and producing a default
plot with recession shading.

The chunks below require an API key (set via `fred_set_key()` or the
`FRED_API_KEY` environment variable). If no key is set, code is shown but
not executed.

## Setup

```{r}
library(fred)
```

## A core macro panel

A typical first move is to grab GDP, unemployment, and headline CPI together,
on a long format that is easy to iterate over.

```{r}
panel <- fred_series(
  c("GDPC1", "UNRATE", "CPIAUCSL"),
  from = "2000-01-01"
)
panel
```

The print header tells you immediately what you are looking at: three series,
how many rows, which transformation was applied, the frequency, and any
vintage information. Underneath, `panel` is a plain data frame.

## Transformations

For inflation and growth analyses, raw levels are usually less useful than
year-on-year percent change. Pass a readable `transform` rather than the raw
FRED `units` code:

```{r}
growth_panel <- fred_series(
  c("GDPC1", "CPIAUCSL"),
  from = "2010-01-01",
  transform = "yoy_pct"
)
growth_panel
```

`transform` and `units` are mutually exclusive. The full list of readable
aliases lives in `?fred_series`. Server-side transforms avoid a manual
`diff(log(x))` step and ensure the result matches what FRED itself shows.

## Wide format for cross-series work

For correlation matrices, scatter plots, or regression input, wide is more
convenient than long.

```{r}
wide <- fred_series(
  c("UNRATE", "CIVPART", "EMRATIO"),
  from = "2000-01-01",
  format = "wide"
)
head(wide)
```

## Default plot

The default `plot.fred_tbl()` method handles long and wide format
automatically and shades NBER recessions when the date range overlaps one.

```{r}
plot(growth_panel, ylab = "% YoY")
```

```{r}
plot(wide, ylab = "%", main = "US labour market")
```

Pass `recessions = FALSE` to switch off shading. Pass a `col =` vector of
length equal to the number of series to override the default palette.

## A starting catalogue

If you cannot remember a series ID, `fred_catalogue()` ships a curated
offline reference of around 50 widely used series:

```{r, eval = TRUE}
fred_catalogue(category = "Inflation")
```

Filtering with `query =` does a case-insensitive substring match against the
ID, title, and description:

```{r, eval = TRUE}
fred_catalogue(query = "mortgage")
```

For the FRED category tree, use `fred_browse()`. With no arguments it shows
the eight top-level categories from a static reference (no API call):

```{r, eval = TRUE}
fred_browse()
```

Pass a category ID to drill into its children (this hits the API and is
cached).

## Where to next

* **Real-time and vintages:** see `vignette("nowcasting-with-fred")` for a
  pseudo-real-time GDP nowcasting backtest.
* **Inflation revisions:** see `vignette("inflation-revisions")` for tracking
  how core inflation estimates change with revisions.
* **Reproducibility:** combine `fred_cite_series()` and `fred_manifest()` to
  pin every series your analysis depends on.
